Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-12-11 manu-draft.qmd
minor text edit

Climate change is an existential threat to environmental resources that sustain human health and the environment. In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth. Changing physical conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be associated climate stress. Three datasets of varying sample design and coverage were evaluated to characterize physical environments in Tampa Bay and the potential relationships with seagrass change. Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 ppt per year. These changes varied spatially and seasonally, with the most dramatic changes being observed in the upper bay segments during the summer months. Simple linear models provided limited evidence that the recent seagrass decline is associated with hotter and fresher conditions. Trends in warming and increased precipitation in the region will continue, further creating suboptimal conditions for seagrasses in Tampa Bay. These results provide a motivating example for resource managers to consider the likelihood of reduced resilience of estuarine resources due to shifting ecological baselines from climate change. Conventional management approaches that focus on limiting nutrient loads should be continued, but the effectiveness may be limited and additional actions will be needed to maintain ecosystem health in the future.

@@

Lastly, our result showing that salinity has decreased in Tampa Bay is generally contrary to expectations for coastal systems consistent with sea-level rise projections [@alarcon2024]. Salinity changes with sea-level rise have been documented to contribute to numerous alterations of subtidal and nearshore habitats [@brinson1995; @white2017]. In southwest Florida, the most common ecological example is migration of tidal wetlands, primary mangroves, upland as porewater salinity and water levels have gradually changed over the past few decades [@borchert2018]. Alteration of salinity regimes for surface and groundwater resources are also well documented. In Florida Bay, for example, widespread decline of *Thalassia testudinium* has been partially attributed to elevated salinity levels beyond the optimal range for the species [@hall2016]. Although drought has been implicated in the hypersaline conditions observed in Florida Bay, sea level rise is expected to further alter salinity dynamics in the region. @dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Greater Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters. In Tampa Bay, the long-term trends of decreasing salinity suggest that the hydrologic load into the system has had a greater influence on salinity regimes than the effects of sea-level rise. This hypothesis is supported by our assessment of precipitation patterns over time, where the long-term increase is inversely correlated with the decrease in salinity.

>>>>

Climate change is an existential threat to environmental resources that sustain human health and the environment. In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth. Changing physical conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress. Three datasets of varying sample design and coverage were evaluated to characterize physical environments in Tampa Bay and the potential relationships with seagrass change. Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 ppt per year. These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay segments during the summer months. Simple linear models provided limited evidence that the recent seagrass decline is associated with hotter and fresher conditions. Trends in warming and increased precipitation in the region will continue, further creating suboptimal conditions for seagrasses in Tampa Bay. These results provide a motivating example for resource managers to consider the likelihood of reduced resilience of estuarine resources due to shifting ecological baselines from climate change. Conventional management approaches that focus on limiting nutrient loads should be continued, but the effectiveness may be limited and additional actions will be needed to maintain ecosystem health in the future.

@@

Lastly, our result showing that salinity has decreased in Tampa Bay is generally contrary to expectations for coastal systems consistent with sea-level rise projections [@alarcon2024]. Salinity changes with sea-level rise have caused numerous alterations of subtidal and nearshore habitats [@brinson1995; @white2017]. In southwest Florida, the most common ecological example is migration of tidal wetlands, primarily mangroves, upland as porewater salinity and water levels have gradually changed over the past few decades [@borchert2018]. Alteration of salinity regimes for surface and groundwater resources are also well documented. In Florida Bay, for example, widespread decline of *Thalassia testudinium* has been partially attributed to elevated salinity levels beyond the optimal range for the species [@hall2016]. Although drought has been implicated in the hypersaline conditions observed in Florida Bay, sea level rise is expected to further alter salinity dynamics in the region. @dessu2018 noted that sea level rise is expected to have the largest effect on salinity changes during periods of low freshwater outflow from the Greater Florida Everglades, emphasizing that measured salinity represents the relative contributions of oceanic and freshwater surface waters. In Tampa Bay, the long-term trends of decreasing salinity suggest that the hydrologic load into the system has had a greater influence on salinity regimes than the effects of sea-level rise. This hypothesis is supported by our assessment of precipitation patterns over time, where the long-term increase is inversely correlated with the decrease in salinity.
